
---
title: Extreme Weather and Terrorism
author: Bradley Druzinsky and Connor Ford
output: html_document
---

Import/install necessary packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(plyr)
library(stringr)
library(sp)
require(plyr)
require(dplyr)
require(tidyr)
library(ggmap)
library(sp)
library(maps)
library(maptools)
```


Import the Terrorism and Temperature data from split csv's:
```{r}
## Aggregating .csv's
read <- function(path) {
  read.csv(path,header=F,sep=",", fileEncoding="latin1")
}

## Load data from .csv's
load_data <- function(path) { 
  files <- dir(path, pattern = '\\.csv', full.names = TRUE)
  tables <- lapply(files, read)
  rbind.fill(tables)
}

## master data tables
terrorismData <- load_data("./TerrorismData")
temperatureData <- load_data("./TemperatureData")

## make first row the header
names(terrorismData) <- as.matrix(terrorismData[1, ])
names(temperatureData) <- as.matrix(temperatureData[1, ])

terrorismData <- terrorismData[-1, ]
temperatureData <- temperatureData[-1, ]

terrorismData[] <- lapply(terrorismData, function(x) type.convert(as.character(x)))
temperatureData[] <- lapply(temperatureData, function(x) type.convert(as.character(x)))
```

View the initial datasets
```{r}
## View head and dimensions of Terrorism Data
head(terrorismData)
dim(terrorismData)

## View head and dimensions of Temperature Data
head(temperatureData)
dim(temperatureData)
```

**Some important features (variables) in the terrorism and temperature data sets**

#### Terrorism Set (From url...)
* Event id
* Date
* Country
* City
* Latitude, Longitude
* Summary (of attack)
* Attack type (hostage, assassination, bombing, etc..)
* Weapon type
* Number killed

#### Temperature Set (From url...)
* Date
* Average Temperature (of day)
* City
* Country
* Latitude, Longitude

#### Lat/Long -> Clustered Regions
The function below takes a vector of Latitudes and a vector of Longitudes and runs Hartigan and Wong's (1979) k-means clustering algorithm to cluster the latitude and longitudes into regions. This saves us the hassle of converting cities or lat/long coordinates to US states/territories. The function is also easily scalable for non-US lat/long coordinates.
```{r}
cluster <- function(latVector, longVector) {

  set.seed(4747)                                                ## Set seed for reproducibility
  
  lat_name <- "realLat"                                         ## Name realLat column
  long_name <- "realLong"                                       ## Name realLong column
  latlong <- data.frame(latVector, longVector)                  ## Initialize new data frame with lat and long params
  names(latlong) <- c(lat_name, long_name)                      ## Apply names to frame
  
  latlongCluster <- kmeans(latlong, 20, nstart = 10)            ## Run k-means clustering with n centroids and nstart CV's
  
  group_lat <- "GroupLat"                                       ## Name GroupLat column
  group_long <- "GroupLong"                                     ## Name GroupLong column
  latlongCenterData <- data.frame(latlongCluster$centers)       ## Initialize data frame with centers
  names(latlongCenterData) <- c(group_lat, group_long)          ## Apply names to frame
  
  latlongCenterData$clusterCenter <- 
    apply(latlongCenterData, 1, paste, collapse = ",")          ## Combine latlongCenterData columns into a comma seperated value
  
  group_name <- "GroupNumber"                                   ## Name GroupNumber column
  resultData <- data.frame(latlongCluster$cluster)              ## Initialize data frame with kmeans clusters (region values)
  names(resultData) <- c(group_name)                            ## Apply names to frame
  
  resultData <- resultData %>%
    mutate(LatLongCenter = 
             latlongCenterData$clusterCenter[GroupNumber])      ## Get the appropriate centroid location for each region number
  
  resultData$cluster <- 
    apply(resultData, 1, paste, collapse = ",")                 ## Create a new column `cluster` columns collapsed together
  
  return(resultData$cluster)                                    ## Return the comma seperated cluster data as a vector
}
```


#### Aggregating Master Data Set
The code chunk below cleans, merges, and computes the master data frame for terrorist and temperature data in the UNITED STATES.
```{r}
## Get and clean only US terrorist attacks
usaTerrorismData <- terrorismData %>%
  filter(country == 217)  %>%                                        ## Filter for country == USA
  unite(date1, iyear,imonth,iday, sep="-") %>%                       ## Unite seperate year, month, and day columns into one (y-m-d)
  mutate(dateLub = ymd(date1)) %>%                                   ## Convert date into 'Lubridate' (YYYY-MM-DD)
  filter(!is.na(dateLub))  %>%                                       ## Filter out observations with missing dates (failed to parse)
  mutate(city_char = str_to_lower(as.character(city))) %>%           ## Create new city column as character type
  mutate(month = month(as.POSIXlt(dateLub, format="%d/%m/%Y"))) %>%  ## Create a new month column
  mutate(year = year(as.POSIXlt(dateLub, format="%d/%m/%Y"))) %>%    ## Create a new year column
  mutate(month_year = paste(month, year))                            ## Create new month, year column (M YYYY)
  
## Uncomment to view head and dimension of cleaned usaTerrorismData
#head(usaTerrorismData)
#dim(usaTerrorismData)


## Get and clean only US temperatures
usaTemperatureData <- temperatureData %>%
  filter(Country == "United States") %>%                             ## Filter for country == USA
  mutate(dateLub = ymd(dt)) %>%                                      ## Convert date into 'Lubridate' (YYYY-MM-DD)
  filter(!is.na(dateLub)) %>%                                        ## Filter out observations with missing dates (failed to parse)
  filter(dateLub >= as.Date("1970-01-01")) %>%                       ## Filter only for data after 1/1/1970 (when Terrorism data begins)
  filter(!is.na(AverageTemperature)) %>%                             ## Remove observations with missing temperatures
  mutate(city_char = str_to_lower(as.character(City))) %>%           ## Create new city column as character type
  mutate(month = month(as.POSIXlt(dateLub, format="%d/%m/%Y"))) %>%  ## Create a new month column
  mutate(year = year(as.POSIXlt(dateLub, format="%d/%m/%Y"))) %>%    ## Create a new year column
  mutate(month_year = paste(month, year))                            ## Create new month, year column (M YYYY)
  
## Uncomment to view head and dimension of cleaned usaTemperatureData
#head(usaTemperatureData)
#dim(usaTemperatureData)


## Merge temperature data with terrorism data by city_char and month_year
## As a result, for every city we should also have every month since 1970 (exlcuding missing data)
## (Important note: There are ~200 unique cities in our temperature data and ~700 in our terrorism data.
## As a result, when merging on cities, we lose about half of our terrorist attacks.)
mergedTempTerrorismData <- left_join(usaTemperatureData, usaTerrorismData, by=c("city_char", "month_year"))

## Uncomment to view head and dimension of mergedTempTerrorismData
#head(mergedTempTerrorismData)
#dim(mergedTempTerrorismData)

  
## Clean mergedTempTerrorismData 
mergedTempTerrorismData <- mergedTempTerrorismData[,c("month_year", "month.x", "year.x",
                                                      "city_char", "AverageTemperature", 
                                                      "Latitude", "Longitude", "nkill", 
                                                      "eventid")] %>%                       ## Select explanatory variables
  mutate(isAttack = as.integer(!is.na(eventid))) %>%                                        ## Create binary isAttack variable
  mutate(nkill = ifelse(is.na(nkill), 0, nkill)) %>%                                        ## Convert NA's to 0's in nkill
  select(-eventid) %>%                                                                      ## Remove eventid variable 
  mutate(realLong = -1* as.numeric(str_sub(Longitude, 1, -2))) %>%                          ## Parse Longitude column
  mutate(realLat = as.numeric(str_sub(Latitude, 1, -2))) %>%                                ## Parse Latitude column
  filter(!is.na(realLat)) %>%                                                               ## Remove missing Latitudes
  filter(!is.na(realLong)) %>%                                                              ## Remove missing Longitudes
  mutate(region = cluster(realLat, realLong)) %>%                                           ## Get comma seperated region cluster (r,lat,long)
  separate(region, sep=",", into=c("Region", "RegionLat", "RegionLong"), convert=TRUE)      ## Seperate comma-seperated data into the 3 columns
  
  

## Uncomment to view head and dimension of cleaned mergedTempTerrorismData
head(mergedTempTerrorismData)
dim(mergedTempTerrorismData)

```








Visualizations:
```{r}





library(plotly)
library(shiny)

terroristAttacks <- usaTerrorismData %>% filter(provstate != "Puerto Rico") %>% filter(provstate != "Alaska") %>% filter(provstate != "Hawaii") %>% filter(longitude < 0)

terroristAttacksByState <- terroristAttacks %>% group_by(provstate) %>% summarise(sum(nkill))


terroristAttacksByState$hover <- with(terroristAttacksByState, paste(provstate, '<br>', "Number Killed", nkill))
# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

p <- plot_geo(terroristAttacksByState, locationmode = 'USA-states') %>%
  add_trace(
    z = ~ nkill, text = ~hover, locations = ~code,
    color = ~nkill, colors = 'Reds'
  ) %>%
  colorbar(title = "Deaths") %>%
  layout(
    title = 'Deaths From Terrorist Attacks Since 1970',
    geo = g
  )


usa <- map_data("usa") # we already did this, but we can do it again
ggplot() + geom_polygon(data = usa, aes (x = long, y = lat), fill = "white") + geom_point(data = terroristAttacks, aes(x=longitude, y = latitude), color = "red") + coord_fixed(1.3)
```


